---
title: "231 Project"
author: "Lily Heidger"
format: 
  html:
    code-fold: show
    toc: true
    embed-resources: true
editor: visual
theme: lumen
execute:
  echo: true
  message: false
  warning: false
---

```{r}
.libPaths("~/rlibs")
library(osmdata)
library(here)
library(dplyr)
library(janitor)
library(sf)
library(mapview)
library(terra)
library(sf)
library(ggplot2)
library(tidycensus)
library(tidyverse)
library(knitr)
library(kableExtra)
```
# Introduction
Describes the data, the research questions, provides any background readers need to understand your project, etc.
## Study Area
San Jose, California 
## Data

## Research Questions
- Do areas with higher urban canopy percentages also have more bike infrastructure?

- Can urban canopy serve as a reliable predictor of bike infrastructure at the street-level?

## Tree Canopy Data

```{r}
sj_boundary <- st_read(here("PSTAT_231_25", "data", "san_jose", "urbanboundary", "San_Jose.shp"))
sf_boundary <- st_read(here("PSTAT_231_25","data", "San_Francisco_and_Oakland", "urbanboundary", "San_Francisco_and_Oakland.shp"))

sj_canopy <- rast(here("PSTAT_231_25", "data", "san_jose", "urbancanopy2022", "San_Jose_canopy2022.tif"))
colnames(sj_canopy)

saveRDS(sj_boundary, here("data", "sj_boundary.rds"))
saveRDS(sf_boundary, here("data", "sf_boundary.rds"))


```

## Osm Data

(ChatGPT used to learn how to use `osmdata`)

```{r}
# Define the area of interest (San Jose)
san_jose <- getbb("San Jose, California")

# Get all streets in San Jose
all_streets <- san_jose %>%
  opq() %>%
  add_osm_feature(key = "highway", 
                  value = c("motorway", "trunk", "primary", "secondary", 
                            "tertiary", "residential", "unclassified", 
                            "service", "living_street")) %>%
  osmdata_sf()

# Extract the lines (street segments)
streets_sf <- all_streets$osm_lines
```

```{r}
# Clean and prepare the data
streets_clean <- streets_sf %>%
  select(osm_id, name, highway, cycleway, `cycleway:left`, `cycleway:right`, 
         bicycle, maxspeed, geometry) %>%
  mutate(
    bike_infra = case_when(
      # 🔵 Bike path
      cycleway == "separate" ~ "bike path",
      
      # 🟢 Buffered lane
      cycleway == "track" | `cycleway:left` == "track" ~ "buffered lane",
      
      # 🟡 Lane
      cycleway == "lane" | 
        `cycleway:left` == "lane" | 
        `cycleway:right` == "lane" | 
        `cycleway:right` == "lane" ~ "lane",
      
      # 🟠 Sharrow
      cycleway == "shared_lane" |
        `cycleway:left` == "shared_lane" |
        `cycleway:right` == "shared_lane" ~ "sharrow",
      
      # ⚪ None
      cycleway %in% c("no", "opposite_lane") |
        `cycleway:left` == "no" |
        `cycleway:right` == "no" ~ "none",
      
      # Default fallback
      TRUE ~ "none"
    )
  )


# For visualization

ggplot(streets_clean, aes(bike_infra))+
  geom_bar()


# ggplot() +
#   geom_sf(data = streets_clean, aes(color = bike_infra), size = 0.8) +
#   scale_color_brewer(palette = "Set1") +
#   theme_minimal() +
#   labs(title = "San Jose Streets with Bike Infrastructure", 
#        color = "Bike Infrastructure Type")

streets_clean %>%
  count(bike_infra) %>%
  arrange(desc(n))


streets_clean <- streets_clean %>%
  mutate(
    maxspeed = as.numeric(gsub("[^0-9.]", "", maxspeed)) 
  ) %>%
  filter(maxspeed <= 60 | is.na(maxspeed)) %>% select(-`cycleway:left`, -`cycleway:right`)

#st_write(streets_clean, "5san_jose_streets_clean.gpkg")

```

```{r}
# Don't think I need this anymore?? 

# Get more comprehensive bike infrastructure information
# streets_detailed <- streets_sf %>%
#   select(osm_id, name, highway, 
#          cycleway, `cycleway:left`, `cycleway:right`, 
#          `cycleway:both`, bicycle, 
#          surface, maxspeed, geometry) %>%
#   # Create a detailed bike infrastructure column
#   mutate(
#     bike_infra_type = case_when(
#       !is.na(cycleway) & cycleway == "lane" ~ "bike lane",
#       !is.na(cycleway) & cycleway == "track" ~ "protected bike lane",
#       !is.na(cycleway) & cycleway == "shared_lane" ~ "sharrow",
#       !is.na(`cycleway:left`) ~ paste("left:", `cycleway:left`),
#       !is.na(`cycleway:right`) ~ paste("right:", `cycleway:right`),
#       !is.na(`cycleway:both`) ~ paste("both sides:", `cycleway:both`),
#       bicycle == "yes" ~ "bicycle allowed",
#       bicycle == "designated" ~ "designated route",
#       TRUE ~ "none"
#     )
#   )
# 
# streets_detailed <- streets_detailed %>%
#   rename(
#     cycleway_left = `cycleway:left`,
#     cycleway_right = `cycleway:right`,
#     cycleway_both = `cycleway:both`
#   ) %>% filter(maxspeed <= 60)
# 
# streets_detailed %>% ggplot(aes(bike_infra_type)) +
#   geom_density() +
#   guides(x =  guide_axis(angle = 90))
# 
# 
# st_write(streets_detailed, "2san_jose_streets_bike_infra.gpkg")
# 
# st_write(streets_detailed, here("san_jose_streets_bike_infra.shp"))
# # Save to file if needed
# #st_write(streets_detailed, here("san_jose_streets_bike_infra.geojson"))
```

## Read in clean tree and bike data

```{r}
sj_buffer <- st_read(here("PSTAT_231_25","data", "City_Limits", "main_Buffer4_Clip.shp"))

sj_city_bound <- st_read(here("PSTAT_231_25","data", "City_Limits.shp"))
# Map canopy density on each buffered street segment
#mapview(sj_buffer["canopy_den"])

saveRDS(sj_buffer, here("PSTAT_231_25","data", "sj_buffer.rds"))
sj_buffer <- readRDS(here("PSTAT_231_25","data", "sj_buffer.rds"))

```
San Jose City Limits accessed from https://gisdata-csj.opendata.arcgis.com/search?tags=boundaries

### EDA
This should contain at least 3 to 5 visualizations and/or tables and their interpretation/discussion. At minimum you should create a univariate visualization of the outcome(s), a bi-variate or multivariate visualization of the relationship(s) between the outcome and select predictors, etc. Part of an EDA involves asking questions about your data and exploring your data to find the answers.

Now let's do some basic box plots to look and see if there is any correlation between canopy density and bike infrastructure. First, we have to make bike infrastructure an ordinal, numeric field.

```{r}
sj_buffer_clean <- sj_buffer %>% mutate(
    comfort_level = case_when(
      bike_infra == "bike path" ~ "5",

      bike_infra == "buffered lane" ~ "4",
      
      bike_infra == "lane" ~ "3",
      
      bike_infra == "sharrow" ~ "2",
      
      bike_infra == "none" ~ "1"))

as.numeric(sj_buffer_clean$comfort_level)


ggplot(sj_buffer_clean, aes(x = factor(comfort_level), y = canopy_den)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    x = "Bike Infrastructure",
    y = "Tree Canopy Density",
    title = "Canopy Density by Bike Infrastructure Comfort Level"
  ) +
  theme_minimal()

```

## Census data

```{r}
options(tigris_use_cache = TRUE)

sj_blocks <- get_decennial(
  geography = "block group",
  variables = "P1_001N",  # Total population
  state = "CA",
  county = "Santa Clara",
  year = 2020,
  geometry = TRUE
)


#st_write(sj_blocks, "santa_clara_census_blocks.gpkg")

sj_city_bound <- st_transform(sj_city_bound, 4326)
sj_blocks <- st_transform(sj_blocks, 4326)


sj_blocks_clip <- st_intersection(sj_blocks, sj_city_bound)
```

```{r}
set.seed(123)  # For reproducibility

sj_sample <- sj_buffer_clean %>%
  dplyr::slice_sample(n = 5000)

sj_sample <- st_transform(sj_sample, 4326)


# 1. Create buffers around your street segments
buffer_distance <- 25  # meters or whatever distance you prefer
streets_buffers <- st_buffer(sj_sample, buffer_distance) 
streets_buffers <- st_transform(streets_buffers, 4326)

# 2. Get census tract data with demographic variables
# If using tidycensus, something like:
ca_tracts <- get_acs(
  geography = "block group",
  variables = c(
    "B19013_001", # Median household income
    "B15003_017", # High school education
    "B15003_022", # College education
    "B08301_001", # Total commuters
    "B08301_019"  # Bicycle commuters
    # Add other variables as needed
  ),
  state = "CA",
  county = "Santa Clara",
  year = 2020,
  geometry = TRUE
)

# 3. Calculate percentages and process your tract data
tract_data <- ca_tracts %>%
  # Process and calculate percentages
  # This will depend on the specific variables you're using
  mutate(
   income_med = "B19013_001",
    educ_high_school_perc = "B15003_022",
    educ_college_perc = "B15003_017",
    bicycle_perc = "B08301_001",
    male_bicycle_perc = "B08301_019"
  )

tract_data <- st_transform(tract_data, st_crs(streets_buffers))

# Convert the columns to numeric
tract_data <- tract_data %>%
  mutate(
    income_med = as.numeric(income_med),
    educ_high_school_perc = as.numeric(educ_high_school_perc),
    educ_college_perc = as.numeric(educ_college_perc),
    bicycle_perc = as.numeric(bicycle_perc),
    male_bicycle_perc = as.numeric(male_bicycle_perc)
  )

# 4. Perform the interpolation
streets_with_demographics <- cbind(
  sj_sample %>% 
    dplyr::select(osm_id, name),
  st_interpolate_aw(
    tract_data[c(
      "income_med",
      "educ_high_school_perc",
      "educ_college_perc",
      "bicycle_perc",
      "male_bicycle_perc"    )],
    streets_buffers,
    extensive = FALSE,  # FALSE for percentages/rates, TRUE for counts
    keep_NA = TRUE
  ) %>% 
    st_drop_geometry()
)
```
[ChatGPT used to determine how to randomly sample my original sf object due to computational limitations.]

## Merging Datasets

```{r}
sj_buffer_clean <- st_transform(sj_buffer_clean, 4326)




streets_blocks_intersection <- st_intersection(sj_buffer, sj_blocks)

# Calculate length of each street segment within each block
streets_blocks_intersection$segment_length <- st_length(streets_blocks_intersection)

# Group by street segment ID and calculate population density
streets_population <- streets_blocks_intersection %>%
  group_by(osm_id) %>%
  summarize(
    total_length = sum(segment_length),
    weighted_population = sum(P1_001N * as.numeric(segment_length) / total_length),
    population_density = weighted_population / total_length
  )

streets_with_population <- streets_clean %>%
  left_join(streets_population, by = "osm_id")
```

\[[Claude LLM](https://claude.ai/) used to help build code for accurately assigning population weighting to street segments\]

# Data Splitting and Cross-Validation 
Describe your process of splitting data into training, test, and/or validation sets. Describe the process of cross-validation. Remember to write for a general audience. Act as if your project will be read by people new to machine learning.

# Model Fitting 
Describe the types of models you fit, their parameter values, and the results

# Model Selection and Performance 
A table and/or graph describing the performance of your best-fitting model on testing data. Describe your best-fitting model however you choose, and the quality of its predictions, etc

# Conclusion
Discusses the outcome(s) of models you fit. Which models performed well, which performed poorly? Were you surprised by model performance? Next steps? General conclusions?
